---
title: "postgres_migration_cleaning"
author: "Jessica Mahone"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(RSQLite)
library(tidyverse)
library(readxl)

# Connect to SQLite
sqlite_con <- dbConnect(RSQLite::SQLite(), "local_news_100_communities.db")

# Read tables
communities <- dbReadTable(sqlite_con, "communities")
outlets <- dbReadTable(sqlite_con, "outlets")
stories <- dbReadTable(sqlite_con, "stories")
```

## Audit

### Outlet Type
```{r outlet_type_check}
outlets %>%
  count(outlet_type, sort = TRUE)
```

Two issues identified: outlet_id 594 was miscoded and should be "Radio Station"; 
outlet_id 145 is a church website, not a news outlet, and will be removed.

### Original and Local
```{r original_local_check}
stories %>%
  count(original, local, sort = TRUE)
```

~50 rows have null values for both original and local. These appear to be a 
data entry gap and will be removed. Less than 0.5% of observations.

### Publication Date
```{r publication_date_check}
stories %>%
  count(publication_date, sort = TRUE)
```

Publication dates are stored as Unix timestamps — a SQLite artifact. Will be 
converted to proper DATE format during migration. 894 null publication dates are 
concentrated in TV and radio outlets, which often do not display dates on web 
content. These are retained as meaningful nulls.

### Critical Information Need
```{r critical_info_need_check}
stories %>%
  count(critical_info_need, sort = TRUE)
```

Controlled vocabulary is clean with nine categories. 269 null values are 
concentrated in broadcast outlets and are retained as meaningful nulls, consistent 
with the pattern seen in publication dates.

### Community ID Nulls in Outlets
```{r community_id_null_check}
outlets %>%
  filter(is.na(community_id)) %>%
  count()
```

21 outlets are missing community_id. These will be recovered using the original 
coder assignment spreadsheet.

---

## Cleaning

```{r}
stories_clean <- stories %>%
  select(-community_id) %>%
  mutate(
    original = original == "Yes",
    local = local == "Yes",
    publication_date = as.Date(as.POSIXct(publication_date, 
                                          origin = "1970-01-01", 
                                          tz = "UTC"))
  ) %>%
  filter(!is.na(original) & !is.na(local))
```

### Fix Outlet Type Errors
```{r fix_outlet_type}
outlets_clean <- outlets %>%
  mutate(outlet_type = case_when(
    outlet_id == 594 ~ "Radio Station",
    TRUE ~ outlet_type
  )) %>%
  filter(outlet_id != 145)
```

### Recover Missing Community IDs

Load and prepare the coder assignment spreadsheet:
```{r load_coders}
coders <- read_xlsx("/data/raw/Coder URL Assignment List_Summer 2017.xlsx") %>%
  filter(!str_detect(community, "DUPLICATE")) %>%
  separate(community, into = c("city", "state"), sep = ",\\s*")
```

Fix Census city name artifact in communities before joining:
```{r fix_census_city_name}
communities_clean <- communities %>%
  mutate(city = str_replace(city, " City \\(Balance\\)", ""))
```

Build community lookup and fill missing community_ids:
```{r community_lookup}
community_lookup <- coders %>%
  left_join(communities_clean %>% select(community_id, city),
            by = "city") %>%
  select(outlet_id, community_id)

outlets_clean <- outlets_clean %>%
  left_join(community_lookup, by = "outlet_id") %>%
  mutate(community_id = coalesce(community_id.x, community_id.y)) %>%
  select(-community_id.x, -community_id.y)
```

Check remaining nulls:
```{r check_remaining_nulls}
outlets_clean %>%
  filter(is.na(community_id)) %>%
  count()
```

One remaining null is outlet_id 360, flagged as a duplicate entry in the coder 
spreadsheet. Removed:
```{r remove_duplicate}
outlets_clean <- outlets_clean %>%
  filter(outlet_id != 360)
```

---

## Validation

### Row Counts
```{r row_counts}
nrow(communities_clean)
nrow(outlets_clean)
```

### No Remaining Null Community IDs
```{r final_null_check}
outlets_clean %>%
  filter(is.na(community_id)) %>%
  count()
```

### Outlet Type Distribution
```{r final_outlet_type}
outlets_clean %>%
  count(outlet_type, sort = TRUE)
```

---

## Export
```{r export}
write_csv(outlets_clean, "/data/processed/outlets_clean.csv")
write_csv(communities_clean, "/data/processed/communities_clean.csv")
write_csv(stories_clean, "/data/processed/stories_clean.csv")
```